Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-06-09 doc/invt.tex
intq.h: logarithmic grid points

\item \label{step:qT}
Compute the optimal $q(T)$ by solving Eq. \eqref{eq:opt_qT}.
\footnote{We applied the Newton-Raphson method on $\ln q(T)$
    with the initial value set to $q(T) = \ln(1+T\,a_0/2)$.
    Other root-finding methods can be more effective\cite{press3rd}.}
%
%Since $m(Q)$ is a decreasing function of $Q$,
%we can start from an interval $[Q_L, Q_R]$
%that satisfies $m(Q_L) > 2/(T\,a_0) > m(Q_R)$
%(e.g. we can choose $Q_L = 0$,
%  and repeatedly double $Q_R$ until
%  the latter inequality is satisfied).
%%
%We can then iteratively refine the interval
%by substituting $Q_M = (Q_L + Q_R)/2$
%for either $Q_L$ and $Q_R$,
%depending on the sign of $m(Q_M) - 2/(T \, a_0)$,
%until the width of the interval is less than
%the desired precision,
%%
%and set $q(T)$ to $Q_M$.

\item \label{step:alpha}
Compute the optimal schedule by
numerically integrating Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.
%
%This can be done on an even grid of $N$ points
%at $q_{[i]} = \frac{i}{N} \, q(T)$
%($i = 0, 1, \dots, N$)
We have used the trapezoidal rule\cite{press3rd}
  on an adaptively defined grid on $[0, q(T)]$ of $N+1$ points
  such that the resulting change of $t_{[i+1]}-t_{[i]}$
  is no less than $T/(2N)$ and no greater than $2T/N$.
%
The curve $q(t)$ is then defined on the grid points,
$(q_{[i]}, t_{[i]})$,
%
and the updating magnitude can be obtained from numerical differentiation:
    $\alpha_i = (1 - e^{-\lambda_1 \, \hat \alpha_i})/\lambda_1$
    and
    $\hat \alpha_i \equiv (q_{[i+1]} - q_{[i-1]}) / (t_{[i+1]} - t_{[i-1]})$,
with
$q_{[ -1]} \equiv q_{[0]} = 0$,
$q_{[N+1]} \equiv q_{[N]} = q(T)$,
$t_{[ -1]} \equiv t_{[0]} = 0$, and
$t_{[N+1]} \equiv t_{[N]} = T$.
%
The resulting values of $(t_{[i]}, \alpha_{[i]})$
can be linear interpolated
to get the updating magnitude, $\alpha(t)$, at any simulation time, $t$.

@@

The above procedure is still somewhat idealized.
%
The preliminary run in step \ref{step:prerun}
needs to be sufficiently long to precisely
estimate the $\Gamma_k$'s and the initial bias, $\epsilon_k$,
which will be used to compute the optimal $q(T)$
and schedule later.

>>>>

\item \label{step:qT}
Compute the optimal $q(T)$ by solving Eq. \eqref{eq:opt_qT}.
\footnote{We solved the equation for $\ln q(T)$ using
    a variant of the Newton-Raphson method
    with the bisection method as fallback.
    The initial value set to $q(T) = \ln(1+T\,a_0/2)$.
    Other general methods can be more effective\cite{press3rd}.}
%
%Since $m(Q)$ is a decreasing function of $Q$,
%we can start from an interval $[Q_L, Q_R]$
%that satisfies $m(Q_L) > 2/(T\,a_0) > m(Q_R)$
%(e.g. we can choose $Q_L = 0$,
%  and repeatedly double $Q_R$ until
%  the latter inequality is satisfied).
%%
%We can then iteratively refine the interval
%by substituting $Q_M = (Q_L + Q_R)/2$
%for either $Q_L$ and $Q_R$,
%depending on the sign of $m(Q_M) - 2/(T \, a_0)$,
%until the width of the interval is less than
%the desired precision,
%%
%and set $q(T)$ to $Q_M$.

\item \label{step:alpha}
Compute the optimal schedule by
numerically integrating Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.\footnote{
  We have used the trapezoidal rule\cite{press3rd}
  on an adaptively defined grid on $[0, q(T)]$ of $N+1$ points
  (we chose $N=20000$)
  such that the resulting change of $\Delta t_{[i]} \equiv t_{[i+1]}-t_{[i]}$
  is no less than $T/(2N)$ and no greater than $2T/N$.
  %
  This requires a few (usually $2$ or $3$) rounds of iterations
  to determine the constant $C$.
  %
  The first guess of $C$ is obtained from integrating
  Eq. \eqref{eq:mint} on a uniform grid.
  %
  Then in each of the next few rounds, we determine the grid points
  as follows.  For grid point $i = 0, 1, \dots, N - 1$,
  we determine the stride
  $\Delta q_{[i]} \equiv q_{[i+1]} - q_{[i]}$
  by a few iterations.
  Starting with the guess, $\bigl(q(T)-q_{[i]}\bigr)/(N-i)$,
  if the resulting increment $\Delta t_{[i]}$
  for grid point $i$ is less than $T/(2N)$,
  we multiply $\Delta q_{[i]}$ by $3/2$
  unless the resulting $q_{[i+1]} > q(T)$,
  or if $\Delta t_{[i]} > 2T/N$,
  we divide $\Delta q_{[i]}$ by $2/3$.
  %
  Once the grid points are defined,
  we will determine the new value of $C$,
  and if the new value differs from the old estimate
  by more than $10\%$, we start another round of iteration.
%
%The curve $q(t)$ is then defined on the grid points,
%$(q_{[i]}, t_{[i]})$,
%
The updating magnitude can then be obtained from numerical differentiation:
    $\alpha_i = (1 - e^{-\lambda_1 \, \hat \alpha_i})/\lambda_1$
    and
    $\hat \alpha_i \equiv (q_{[i+1]} - q_{[i-1]}) / (t_{[i+1]} - t_{[i-1]})$,
with
$q_{[ -1]} \equiv 0$,
$q_{[N+1]} \equiv q(T)$,
$t_{[ -1]} \equiv 0$, and
$t_{[N+1]} \equiv T$.
%
The resulting values of $(t_{[i]}, \alpha_{[i]})$
can be linear interpolated
to get the updating magnitude, $\alpha(t)$, at any simulation time, $t$.
}

@@

The above procedure is still somewhat idealized.
%
The preliminary run in step \ref{step:prerun}
needs to be sufficiently long to precisely
estimate the $\Gamma_k$'s and the initial bias, $\epsilon_k$
(for which one may use the histogram-corrected bias potential).
%
However,
